---
title: "PA1"
author: "Freddy Li"
output: html_document
---

## Loading and preprocessing the data  
```{r,cache=TRUE, echo=TRUE}
if(!file.exists("activity.csv")){
  url <- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip"
  temp <- tempfile()
  download.file(url,temp)
  unzip(temp)
  unlink(temp)
}
data <- read.csv("activity.csv",stringsAsFactors = FALSE)
str(data)
summary(data)
```

Transform the data into suitable format  
```{r, cache=TRUE, echo=TRUE}
library(lubridate)
library(dplyr)
data$date <- ymd(data$date)
data <- tbl_df(data)
```

## What is the average daily activity pattern?  
#### 1. Calculate the total number of steps taken per day  
```{r, cache=TRUE, echo=TRUE}
total_perday <- aggregate(steps~date,data,sum,na.rm=TRUE)
```

#### 2. Make a histogram of the total number of steps taken each day  
```{r, cache=TRUE, echo=TRUE}
with(total_perday,hist(steps,col="blue",xlab="Numbers of Steps",main = "Total Steps Taken Each Day"))
```     

#### 3. Calculate and report the mean and median of the total number of steps taken per day  
```{r, cache=TRUE, echo=TRUE}
mean_perday <- mean(total_perday$steps)
median_perday <- median(total_perday$steps)
mean_perday
median_perday
```
The mean and median of the total number of steps taken per day are `r mean_perday` and `r median_perday` respectively.  

## What is the average daily activity pattern?  
#### 1. Make a time series plot of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all days (y-axis)  
```{r, cache=TRUE, echo=TRUE}
steps_interval <- aggregate(steps~interval, data, mean, na.rm=TRUE)
plot(steps_interval$interval, steps_interval$steps, type="l",col="blue",lwd=1,xlab="Interval",ylab="Number of Steps",main="Average Number of Steps for given intervals across all days")
```


#### 2. Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?  
```{r, cache=TRUE, echo=TRUE}
max_interval <-steps_interval$interval[max(steps_interval$steps)] 
```
On average across all the days in the dataset, the 5-minute interval, `r max_interval`, contains the maximum number of steps.  

## Imputing missing values  
#### 1. Calculate and report the total number of missing values in the dataset  
```{r, cache=TRUE, echo=TRUE}
na_sum <- sum(is.na(data$steps))
na_percent <- round(mean(is.na(data$steps))*100,2)
```
The total number of missing values is `r na_sum`, which is `r na_percent`% of total observations.    

#### 2. Devise a strategy for filling in all of the missing values in the dataset.  
For the missing values, the mean for that 5-minute interval will be used to fill in.  

#### 3. Create a new dataset that is equal to the original dataset but with the missing data filled in.  
```{r, cache=TRUE, echo=TRUE}
updated_data <-data
updated_data$steps <- ifelse(!is.na(data$steps),data$steps,steps_interval$steps[match(data$interval,steps_interval$interval)])
```

#### 4. Make a histogram of the total number of steps taken each day and Calculate and report the mean and median total number of steps taken per day.  
```{r, cache=TRUE, echo=TRUE}
total_new <- aggregate(steps~date,updated_data,sum)
with(total_new,hist(steps,col="red",xlab="Numbers of Steps",main = "Total Steps Taken Each Day"))
mean_new <- mean(total_new$steps)
median_new <- median(total_new$steps)
diff_mean <- mean_new - mean_perday
diff_median <- round(median_new - median_perday,2)
```

The mean and median of the total number of steps taken per day after filling missing values are `r mean_new`, which is same with the old mean due to difference of `r diff_mean`, and `r median_new`,which is higher than the old median due to difference of `r diff_median`, respectively. The imputed value changes the median of total daily steps a little.  

## Are there differences in activity patterns between weekdays and weekends?  
#### 1. Create a new factor variable in the dataset with two levels – “weekday” and “weekend” indicating whether a given date is a weekday or weekend day.
```{r, cache=TRUE, echo=TRUE}
updated_data$weekday <- factor(ifelse(weekdays(updated_data$date) %in% c("Saturday","Sunday"),"Weekend","Weekday"),levels = c("Weekend","Weekday"))
```

#### 2. Make a panel plot containing a time series plot of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all weekday days or weekend days (y-axis).
```{r,cache=TRUE, echo=TRUE}
library(lattice)
steps_interval_new <- aggregate(steps~interval+weekday,updated_data,mean)
xyplot(steps_interval_new$steps~steps_interval_new$interval|steps_interval_new$weekday,type="l",xlab="Interval",ylab="Number of steps",layout=c(1,2))
```
